The CrypTool Script: \\ Cryptography, Mathematics and More \\ \vspace{30pt} {\large Background reading\\%*[-8pt] for the free e-learning program CrypTool\\ (10th edition -- distributed with CrypTool version 1.4.30)} \vspace{60pt}

### 2009-07-15 script/en/chapters/paper_and_pencil.tex
CT script: Spellcheck in order to have consistent American English

% $Id$
% ..............................................................................
%                V E R S C H L U E S S E L U N G S V E R F A H R E N
%
% Writing rule: Capitel header with all nouns/verbs-words capitalized,
%               All other headers in the normal lower/upper case manner
%               like sentences (without a dot at the end).
%
% ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
% ``primenet''
\newpage
%------------------------------------------------------------------------------
% First Editor: Christine Stötzel, April 2004
% Update and corrections: B. Esslinger, June 2005
% corrected by C. Esslinger, June 2005 
%------------------------------------------------------------------------------
\hypertarget{Kapitel_PaperandPencil}{}

@@

Methods to be successful and wide-spread had to fulfill some attributes which are equally required for modern algorithms:
\begin{itemize}
\item Exhaustive description, almost standardisation (including special cases,
      padding, etc.).
\item Good balance between security and usability 
      (because methods being too complicated were error-prone or
      unacceptably slow).
\end{itemize}

@@

\item {\bf Railfence}\footnote{In CrypTool\index{CrypTool} 
   you can simulate this method under the menu {\bf Crypt/Decrypt \textbackslash{} 
   Symmetric (classic) \textbackslash{} Permutation}: for a railfence with
   2 lines use as key ``B,A'' and accept the default settings (only one
   permutation, where your input is done line-by-line and the output is
   taken column-by-column). 
   Using the key ``A,B'' would start the zigzag pattern below in the
   way, that the first letter is written into the first line instead of the
   second line.}
   \cite{Singh2001}\index{Railfence cipher}:
   The characters of a message are alternately written in two (or more) lines,
   creating a zigzag pattern. The resulting ciphertext is read out 
   line by line.\\
   This is more a children's method.

@@

\begin{table}[h]
\begin{center}
\begin{tabular}{r@{\:}r@{\:}r@{\:}r@{\:}r@{\:}r@{\:}r@{\:}r@{\:}r@{\:}r@{\:}r@{\:}r@{\:}r@{\:}r@{\:}r@{\:}r@{\:}r@{\:}r@{\:}r@{\:}r@{\:}r@{\:}r@{\:}r@{\:}r@{\:}}
	  & n &   & x &   & m &   & l &   & o &   & t &   & a &   & s &   & o &   & i &   & i &   & n \\
	a &   & e &   & a &   & p &   & e &   & f &   & r &   & n &   & p &   & s &   & t &   & o &   \\
\end{tabular}
\caption{Railfence cipher}
\end{center} 
\end{table}

@@

%\newpage % Damit die Tables immer direkt nach dem zugehörigen Item. 
\item {\bf Cadenus} \cite{ACA2002}\index{Cadenus}: Cadenus is a form of
   columnar transposition that uses two keywords.\\
   The 1st keyword is used to swap columns.\\
   The 2nd keyword is used to define the initial letter of each column:
   this 2nd keyword is a permutation of the used alphabet. This permutation
   is written on the left of the first column.
   Afterwards, each column is moved (wrap-around) so that it begins with the
   letter, which is in the the same line as the key letter of the first 
   keyword within the second keyword.\\
   Ciphertext is read out line by line.

@@

   \begin{enumerate}
	\item If both letters can be found in the same column, they are 
           replaced by the letters underneath.
	\item If both letters can be found in the same row, take the letters
           to their right.
	\item If both letters of the digraph are in different columns and rows,
           the replacement letters are obtained by scanning along the row of 
           the first letter up to the column where the other letter occurs 
           and vice versa.
	\item Double letters are treated by special rules, if they appear in one
 	   diagraph. They can be separated by a filler, for example.
   \end{enumerate}

@@

\item {\bf ADFG(V)X}\footnote{%
   In CrypTool\index{CrypTool} you can call this method under the menu
   {\bf Crypt/Decrypt \textbackslash{} Symmetric (classic) \textbackslash{} 
   ADFGVX}.}
   \cite{Singh2001}\index{ADFGVX}%
   : 
   ADFG(V)X-encryption was developed in Germany during WW1. The alphabet 
   is filled into a 5x5- or 6x6-matrix, and columns and rows are marked with
   the letters ADFGX and V, depending on the size of the grille. Each 
   plaintext character is substituted by the corresponding pair of letters. 
   Finally, a (row-) transposition cipher is performed on the resulting text.

@@

   \begin{itemize}
      \item {\bf Bifid/Polybius square/checkerboard} \cite{Goebel2003}: 
         Bifid encryption is the basic form of fractionation. A 5x5-matrix 
         is filled with the plaintext alphabet (see Playfair encryption), 
         rows and columns are numbered, so that each cleartext character can be
         substituted by a pair of digits. Mostly the plaintext is divided into 
         blocks of equal length. The length of blocks (here 5) is another 
         configuration parameter of this cipher. Block-by-block all line
         numbers are read out first, followed by all numbers naming the
         columns.
	 To obtain the ciphertext, the digits are pairwise transformed into
	 letters again. The numbers can be any permutation of (1,2,3,4,5),
         which is one key of configuration parameter of this cipher. Instead
         of numbering rows and columns, a keyword can be used, too.

@@

\newpage % Damit die Tables immer direkt nach dem zugehörigen Item. 
\item {\bf Nicodemus} \cite{ACA2002}: First of all, a simple columnar 
   transposition is carried out. Before reading out the columns, the message
   is encrypted aditionally by Vigen\`ere (all letters of a column are
   enciphered with the corresponding keyword letter). The ciphertext is read
   out in vertical blocks.

>>>>

% $Id$
% ..............................................................................
%                V E R S C H L U E S S E L U N G S V E R F A H R E N
%
% Writing rule: Capitel header with all nouns/verbs-words capitalized,
%               All other headers in the normal lower/upper case manner
%               like sentences (without a dot at the end).
% Writing rule: Use American English instead of British English consistently.
%
% ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
% ``primenet''
\newpage
%------------------------------------------------------------------------------
% First Editor: Christine Stötzel, April 2004
% Update and corrections: B. Esslinger, June 2005
% corrected by C. Esslinger, June 2005 
%------------------------------------------------------------------------------
\hypertarget{Kapitel_PaperandPencil}{}

@@

Methods to be successful and wide-spread had to fulfill some attributes which are equally required for modern algorithms:
\begin{itemize}
\item Exhaustive description, almost standardization (including special cases,
      padding, etc.).
\item Good balance between security and usability 
      (because methods being too complicated were error-prone or
      unacceptably slow).
\end{itemize}

@@

\item {\bf Rail Fence}\footnote{In CrypTool\index{CrypTool} 
   you can simulate this method under the menu {\bf Crypt/Decrypt \textbackslash{} 
   Symmetric (classic) \textbackslash{} Permutation}: for a railfence with
   2 lines use as key ``B,A'' and accept the default settings (only one
   permutation, where your input is done line-by-line and the output is
   taken column-by-column). 
   Using the key ``A,B'' would start the zigzag pattern below in the
   way, that the first letter is written into the first line instead of the
   second line.}
   \cite{Singh2001}\index{Rail Fence cipher}:
   The characters of a message are alternately written in two (or more) lines,
   creating a zigzag pattern. The resulting ciphertext is read out 
   line by line.\\
   This is more a children's method.

@@

\begin{table}[h]
\begin{center}
\begin{tabular}{r@{\:}r@{\:}r@{\:}r@{\:}r@{\:}r@{\:}r@{\:}r@{\:}r@{\:}r@{\:}r@{\:}r@{\:}r@{\:}r@{\:}r@{\:}r@{\:}r@{\:}r@{\:}r@{\:}r@{\:}r@{\:}r@{\:}r@{\:}r@{\:}}
	  & n &   & x &   & m &   & l &   & o &   & t &   & a &   & s &   & o &   & i &   & i &   & n \\
	a &   & e &   & a &   & p &   & e &   & f &   & r &   & n &   & p &   & s &   & t &   & o &   \\
\end{tabular}
\caption{Rail Fence cipher}
\end{center} 
\end{table}

@@

%\newpage % Damit die Tables immer direkt nach dem zugehörigen Item. 
\item {\bf Cadenus} \cite{ACA2002}\index{Cadenus}: Cadenus is a form of
   columnar transposition that uses two keywords.\\
   The 1st keyword is used to swap columns.\\
   The 2nd keyword is used to define the initial letter of each column:
   this 2nd keyword is a permutation of the used alphabet. This permutation
   is written on the left of the first column.
   Afterwards, each column is moved (wrap-around) so that it begins with the
   letter, which is in the same line as the key letter of the first 
   keyword within the second keyword.\\
   Ciphertext is read out line by line.

@@

   \begin{enumerate}
	\item If both letters can be found in the same column, they are 
           replaced by the letters underneath.
	\item If both letters can be found in the same row, take the letters
           to their right.
	\item If both letters of the digraph are in different columns and rows,
           the replacement letters are obtained by scanning along the row of 
           the first letter up to the column where the other letter occurs 
           and vice versa.
	\item Double letters are treated by special rules, if they appear in one
 	   digraph. They can be separated by a filler, for example.
   \end{enumerate}

@@

\item {\bf ADFG(V)X}\footnote{%
   In CrypTool\index{CrypTool} you can call this method under the menu
   {\bf Crypt/Decrypt \textbackslash{} Symmetric (classic) \textbackslash{} 
   ADFGVX}.}
   \cite{Singh2001}\index{ADFGVX}%
   : 
   ADFG(V)X-encryption was developed in Germany during WW1. The alphabet 
   is filled into a 5x5 or 6x6 matrix, and columns and rows are marked with
   the letters ADFGX and V, depending on the size of the grille. Each 
   plaintext character is substituted by the corresponding pair of letters. 
   Finally, a (row-) transposition cipher is performed on the resulting text.

@@

   \begin{itemize}
      \item {\bf Bifid/Polybius square/checkerboard} \cite{Goebel2003}: 
         Bifid encryption is the basic form of fractionation. A 5x5 matrix 
         is filled with the plaintext alphabet (see Playfair encryption), 
         rows and columns are numbered, so that each cleartext character can be
         substituted by a pair of digits. Mostly the plaintext is divided into 
         blocks of equal length. The length of blocks (here 5) is another 
         configuration parameter of this cipher. Block-by-block all line
         numbers are read out first, followed by all numbers naming the
         columns.
	 To obtain the ciphertext, the digits are pairwise transformed into
	 letters again. The numbers can be any permutation of (1,2,3,4,5),
         which is one key of configuration parameter of this cipher. Instead
         of numbering rows and columns, a keyword can be used, too.

@@

\newpage % Damit die Tables immer direkt nach dem zugehörigen Item. 
\item {\bf Nicodemus} \cite{ACA2002}: First of all, a simple columnar 
   transposition is carried out. Before reading out the columns, the message
   is encrypted additionally by Vigen\`ere (all letters of a column are
   enciphered with the corresponding keyword letter). The ciphertext is read
   out in vertical blocks.

### 2009-07-15 script/en/chapters/primes.tex
CT script: Spellcheck in order to have consistent American English

The sequence of prime numbers starts with $$ 2,~ 3,~ 5,~ 7, ~ 11, ~ 13, ~
17, ~ 19, ~ 23, ~ 29, ~ 31, ~ 37, ~ 41, ~ 43, ~ 47, ~ 53, ~ 59, ~ 61,
~ 67, ~ 71, ~ 73, ~ 79, ~ 83, ~ 89, ~ 97, \cdots . $$
The first 100 numbers include precisely 25 prime numbers. After this,
the percentage of primes constantly decreases. Prime numbers can be
factorised in a uniquely {\em trivial} way: 
$$5 = 1 \cdot 5,\quad  17 = 1 \cdot 17, \quad 1,013 = 1 \cdot 1,013,  \quad
1,296,409 = 1 \cdot 1,296,409.$$
All numbers that have $2$ or more factors not equal 1 are called 
\index{Number!composite} {\em composite} numbers. 
These include $$ 4 = 2 \cdot 2, \quad 6 = 2\cdot 3 $$ as well
as numbers that {\em look like primes}, but are in fact composite:
$$ 91 = 7 \cdot 13, \quad 161=7 \cdot 23, \quad 767 =13 \cdot 59. $$

@@

\begin{theorem}\label{thm-pz-prod}
Each element $n$ of the natural numbers greater than $1$ can be written as the
product $n = p_1 \cdot p_2 \dots p_m$ of prime numbers. If two such
factorisations $$n =  p_1 \cdot p_2 \cdot \cdots \cdot p_m = p'_1 \cdot p'_2 \cdots
p'_{m'}$$ are given, then they can be reordered such that $\;m = m'\;$ and for
all $i$:  $\;p_i = p'_i$. \\
($p_1, p_2, \dots, p_m$ are called the prime factors of n).
\end{theorem}

In other words: each natural number other than $1$ can be written as a product
of prime numbers in precisely one way, if we ignore the order of the factors.
The factors are therefore unique (the {\em expression as a product of factors}
is unique)! For example, $$ 60 = 2 \cdot 2 \cdot 3 \cdot 5 = 2^2\cdot 3^1 \cdot
5^1. $$
And this --- other than changing the order of the factors --- is the only way in
which the number $60$ can be factorised. If you allow numbers other than primes
as factors, there are several ways of factorising integers and the \textbf{uniqueness} \hypertarget{uniqueness}{} is
lost: $$ 60 = 1 \cdot 60 = 2 \cdot 30 = 4 \cdot 15 = 5 \cdot 12 =6 \cdot 10 = 2
\cdot 3 \cdot 10 = 2 \cdot 5 \cdot 6 = 3 \cdot 4 \cdot 5 = \cdots . $$

@@

\begin{Proof}{}
The theorem of Mersenne can be proved by contradiction%
\index{Proof by contradiction}.  We therefore assume that
there exists a composite natural number $ n $ (with real factorisation)  
$ n=n_1 \cdot n_2 $
, with the property that $ 2^n -1 $ is a prime number.

@@

Mersenne claimed that $2^{67}-1$ is a prime number. There is also a mathematical
history behind this claim: it first took over 200 years before 
\index{Lucas, Edouard} Edouard Lucas (1842-1891) proved that this number 
is composite.
However, he argued indirectly and did not name any of the factors. Then Frank
Nelson Cole\index{Cole, Frank Nelson}\footnote{%
Frank Nelson Cole, American mathematician, Sep. 20, 1861 $-$ May 26, 1926.}
showed in 1903 which factors make up this composite number: 
$$ 2^{67} -1
=147, 573, 952, 589, 676, 412, 927 = 193, 707, 721 \cdot 761, 838, 257, 287. $$
He admitted to having worked 20 years on the factorisation 
\index{Factorisation} (expression as a product of prime factors)\footnote{%
  Using CrypTool\index{CrypTool} you can factorize numbers in the 
  following way: menu {\bf Indiv. Procedures \textbackslash{} RSA Cryptosystem 
  \textbackslash{} Factorisation of a Number}. \\
  CrypTool can factorize in a reasonable time numbers no longer than 250 bit.
  Numbers bigger than 1024 bits are currently not accepted by CrypTool. \\
  The current factorization records are listed in chapter \ref{NoteFactorisation}.
  \index{Factorisation!factoring records}
}
of this 21-digit decimal number!

@@

% --------------------------------------------------------------------------
\vskip +25 pt
\subsection{Challenge of the Electronic Frontier Foundation (EFF)}\index{EFF}
This search is also spurred on by a competition started by the non-profit
organisation EFF (Electronic Frontier Foundation) using the means of an 
unknown donor. The participants are rewarded with a total of 500,000 USD if
they find the longest prime number. In promoting this project, the unknown
donor is not looking for the quickest computer, but rather wants to draw 
people's attention to the opportunities offered by {\em cooperative networking} \\
{\href{http://www.eff.org/awards/coop}{\tt http://www.eff.org/awards/coop}}

@@

If we look for the prime factors in order to decide whether a number is prime,
then the search takes too long, if even the smallest prime factor is enormous.
Factorising numbers using systematic computational
division or using the \hyperlink{SieveEratosthenes01}{sieve of Eratosthenes} 
\index{Eratosthenes!sieve} is only feasible using current computers for
numbers with up to around $20$ digits in the decimal system.
The biggest number factorized into its 2 almost equal prime factors 
has 200 digits 
(see \hyperlink{RSA-200-chap3}{RSA-200} in chapter \ref{NoteFactorisation}).
% be_2005_UPDATEN_if-new-factorization-record-appears

@@

As early as in the 19th century, it was discovered that the $29$-digit number $$
f(7) = 2^{2^7} + 1 $$ is not prime. However, it was not until 1970 that
Morrison/Billhart managed to factorise it.
\begin{eqnarray*}\label{F7Morrison}
f(7) & = & 340,282,366,920,938,463,463,374,607,431,768,211,457 \\
& = & 59, 649, 589, 127, 497, 217 \cdot  5,704,689,200,685,129,054,721
\end{eqnarray*}

@@

   This first generalisation of the Mersenne
   numbers\index{Mersenne!number!generalized}
   creates the so called Proth numbers\index{Number!Proth}.
   There are (for small $k$) extremely quick prime number tests
   (see \cite{Knuth1981}). This can be performed in practice using software
   such as the Proths software from Yves Gallot\index{Gallot, Yves}:
\vspace{-10pt}
\begin{itemize}
  \item[] {\href{http://www.prothsearch.net/index.html}
            {\tt http://www.prothsearch.net/index.html}}.
\end{itemize}

@@

You can now find a large number of finished programs on the Internet - often
complete with source code - allowing you to experiment with large numbers
yourself (see chapter \ref{spezialzahlentypen}). 
You also have access to large databases that contain either a large
number of primes or the factorisation of numerous composite numbers.

@@

      \item Does a polynomial time algorithm exist that calculates for a
            composite number $n$ a non-trivial (i.e. other than $1$ and $n$)
	    factor of $n$?\footnote{Please compare chapters \ref{RSABernstein}
	    and \ref{NoteFactorisation}.}
   \end{itemize}
\end{itemize}

At the end of chapter \ref{NoteFactorisation},
section \hyperlink{RSA-200-chap3}{RSA-200}
you can see the dimensions of the numbers where the current algorithms
testing for primality\index{Primality testing} and
calculating the factorization\index{Factorisation!factoring records}
deliver results.

@@

\bibitem[Aaronson2003]{Aaronson2003} \index{Aaronson 2003}
    Scott Aaronson, \\
    {\em The Prime Facts: From Euclid to AKS}, \\
    \href{http://www.scottaaronson.com/writings/prime.pdf}
 {\texttt{http://www.scottaaronson.com/writings/prime.pdf}}. \\
    After I had completed this article, I did come across the 
    fine paper by Scott Aaronson, which also offers
    a didactically very well-done intoduction to this topic. It is 
    humorous and easy to read but at the same time precise and erudite.

>>>>

The sequence of prime numbers starts with $$ 2,~ 3,~ 5,~ 7, ~ 11, ~ 13, ~
17, ~ 19, ~ 23, ~ 29, ~ 31, ~ 37, ~ 41, ~ 43, ~ 47, ~ 53, ~ 59, ~ 61,
~ 67, ~ 71, ~ 73, ~ 79, ~ 83, ~ 89, ~ 97, \cdots . $$
The first 100 numbers include precisely 25 prime numbers. After this,
the percentage of primes constantly decreases. Prime numbers can be
factorized in a uniquely {\em trivial} way: 
$$5 = 1 \cdot 5,\quad  17 = 1 \cdot 17, \quad 1,013 = 1 \cdot 1,013,  \quad
1,296,409 = 1 \cdot 1,296,409.$$
All numbers that have $2$ or more factors not equal 1 are called 
\index{Number!composite} {\em composite} numbers. 
These include $$ 4 = 2 \cdot 2, \quad 6 = 2\cdot 3 $$ as well
as numbers that {\em look like primes}, but are in fact composite:
$$ 91 = 7 \cdot 13, \quad 161=7 \cdot 23, \quad 767 =13 \cdot 59. $$

@@

\begin{theorem}\label{thm-pz-prod}
Each element $n$ of the natural numbers greater than $1$ can be written as the
product $n = p_1 \cdot p_2 \dots p_m$ of prime numbers. If two such
factorizations $$n =  p_1 \cdot p_2 \cdot \cdots \cdot p_m = p'_1 \cdot p'_2 \cdots
p'_{m'}$$ are given, then they can be reordered such that $\;m = m'\;$ and for
all $i$:  $\;p_i = p'_i$. \\
($p_1, p_2, \dots, p_m$ are called the prime factors of n).
\end{theorem}

In other words: each natural number other than $1$ can be written as a product
of prime numbers in precisely one way, if we ignore the order of the factors.
The factors are therefore unique (the {\em expression as a product of factors}
is unique)! For example, $$ 60 = 2 \cdot 2 \cdot 3 \cdot 5 = 2^2\cdot 3^1 \cdot
5^1. $$
And this --- other than changing the order of the factors --- is the only way in
which the number $60$ can be factorized. If you allow numbers other than primes
as factors, there are several ways of factorizing integers and the \textbf{uniqueness} \hypertarget{uniqueness}{} is
lost: $$ 60 = 1 \cdot 60 = 2 \cdot 30 = 4 \cdot 15 = 5 \cdot 12 =6 \cdot 10 = 2
\cdot 3 \cdot 10 = 2 \cdot 5 \cdot 6 = 3 \cdot 4 \cdot 5 = \cdots . $$

@@

\begin{Proof}{}
The theorem of Mersenne can be proved by contradiction%
\index{Proof by contradiction}.  We therefore assume that
there exists a composite natural number $ n $ (with real factorization)  
$ n=n_1 \cdot n_2 $
, with the property that $ 2^n -1 $ is a prime number.

@@

Mersenne claimed that $2^{67}-1$ is a prime number. There is also a mathematical
history behind this claim: it first took over 200 years before 
\index{Lucas, Edouard} Edouard Lucas (1842-1891) proved that this number 
is composite.
However, he argued indirectly and did not name any of the factors. Then Frank
Nelson Cole\index{Cole, Frank Nelson}\footnote{%
Frank Nelson Cole, American mathematician, Sep. 20, 1861 $-$ May 26, 1926.}
showed in 1903 which factors make up this composite number: 
$$ 2^{67} -1
=147, 573, 952, 589, 676, 412, 927 = 193, 707, 721 \cdot 761, 838, 257, 287. $$
He admitted to having worked 20 years on the factorization 
\index{Factorization} (expression as a product of prime factors)\footnote{%
  Using CrypTool\index{CrypTool} you can factorize numbers in the 
  following way: menu {\bf Indiv. Procedures \textbackslash{} RSA Cryptosystem 
  \textbackslash{} Factorization of a Number}. \\
  CrypTool can factorize in a reasonable time numbers no longer than 250 bit.
  Numbers bigger than 1024 bits are currently not accepted by CrypTool. \\
  The current factorization records are listed in chapter \ref{NoteFactorization}.
  \index{Factorization!factoring records}
}
of this 21-digit decimal number!

@@

% --------------------------------------------------------------------------
\vskip +25 pt
\subsection{Challenge of the Electronic Frontier Foundation (EFF)}\index{EFF}
This search is also spurred on by a competition started by the non-profit
organization EFF (Electronic Frontier Foundation) using the means of an 
unknown donor. The participants are rewarded with a total of 500,000 USD if
they find the longest prime number. In promoting this project, the unknown
donor is not looking for the quickest computer, but rather wants to draw 
people's attention to the opportunities offered by {\em cooperative networking} \\
{\href{http://www.eff.org/awards/coop}{\tt http://www.eff.org/awards/coop}}

@@

If we look for the prime factors in order to decide whether a number is prime,
then the search takes too long, if even the smallest prime factor is enormous.
Factorizing numbers using systematic computational
division or using the \hyperlink{SieveEratosthenes01}{sieve of Eratosthenes} 
\index{Eratosthenes!sieve} is only feasible using current computers for
numbers with up to around $20$ digits in the decimal system.
The biggest number factorized into its 2 almost equal prime factors 
has 200 digits 
(see \hyperlink{RSA-200-chap3}{RSA-200} in chapter \ref{NoteFactorization}).
% be_2005_UPDATEN_if-new-factorization-record-appears

@@

As early as in the 19th century, it was discovered that the $29$-digit number $$
f(7) = 2^{2^7} + 1 $$ is not prime. However, it was not until 1970 that
Morrison/Billhart managed to factorize it.
\begin{eqnarray*}\label{F7Morrison}
f(7) & = & 340,282,366,920,938,463,463,374,607,431,768,211,457 \\
& = & 59, 649, 589, 127, 497, 217 \cdot  5,704,689,200,685,129,054,721
\end{eqnarray*}

@@

   This first generalization of the Mersenne
   numbers\index{Mersenne!number!generalized}
   creates the so called Proth numbers\index{Number!Proth}.
   There are (for small $k$) extremely quick prime number tests
   (see \cite{Knuth1981}). This can be performed in practice using software
   such as the Proths software from Yves Gallot\index{Gallot, Yves}:
\vspace{-10pt}
\begin{itemize}
  \item[] {\href{http://www.prothsearch.net/index.html}
            {\tt http://www.prothsearch.net/index.html}}.
\end{itemize}

@@

You can now find a large number of finished programs on the Internet - often
complete with source code - allowing you to experiment with large numbers
yourself (see chapter \ref{spezialzahlentypen}). 
You also have access to large databases that contain either a large
number of primes or the factorization of numerous composite numbers.

@@

      \item Does a polynomial time algorithm exist that calculates for a
            composite number $n$ a non-trivial (i.e. other than $1$ and $n$)
	    factor of $n$?\footnote{Please compare chapters \ref{RSABernstein}
	    and \ref{NoteFactorization}.}
   \end{itemize}
\end{itemize}

At the end of chapter \ref{NoteFactorization},
section \hyperlink{RSA-200-chap3}{RSA-200}
you can see the dimensions of the numbers where the current algorithms
testing for primality\index{Primality testing} and
calculating the factorization\index{Factorization!factoring records}
deliver results.

@@

\bibitem[Aaronson2003]{Aaronson2003} \index{Aaronson 2003}
    Scott Aaronson, \\
    {\em The Prime Facts: From Euclid to AKS}, \\
    \href{http://www.scottaaronson.com/writings/prime.pdf}
 {\texttt{http://www.scottaaronson.com/writings/prime.pdf}}. \\
    After I had completed this article, I did come across the 
    fine paper by Scott Aaronson, which also offers
    a didactically very well-done introduction to this topic. It is 
    humorous and easy to read but at the same time precise and erudite.

### 2009-07-15 script/en/script-en.tex
CT script: Spellcheck in order to have consistent American English

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\title{ 
	The CrypTool Script: \\
	Cryptography, Mathematics and More \\
        \vspace{30pt}
        {\large	Background reading\\%*[-8pt] 
        for the free eLearning program CrypTool\\
       (10th edition -- distributed with CrypTool version 1.4.30)}
        \vspace{60pt}
}

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\title{ 
	The CrypTool Script: \\
	Cryptography, Mathematics and More \\
        \vspace{30pt}
        {\large	Background reading\\%*[-8pt] 
        for the free e-learning program CrypTool\\
       (10th edition -- distributed with CrypTool version 1.4.30)}
        \vspace{60pt}
}
